Title Text

### 2015-03-13 docs/paper.tex
More text

\documentclass{sigplanconf}
\usepackage{amsmath}
\usepackage{fancyvrb}
\usepackage{url}

@@

\begin{abstract}
We present a technique for integrating GHC's type-checker with the
functionality provided by an SMT solver.  The technique was developed
to add support for reasoning about type-level functions on natural
numbers, and so our implementation mainly uses the theory of linear
arithmetic.  The technique, however, is more general and makes it
possible to experiment with other external decision procedures.
\end{abstract}

@@

\section {SMT Solvers}

@@

This example asks the solver two questions, and the answer to the first
one is \Verb"unsatisfiable", while the answer to the second one is
\Verb"satisfiable".

The \Verb"push" command instructs to solver to save its state, then
proceed as normal.  When the solver encounters a \Verb"pop" command,
it reverts to the last saved state.

This functionality is extremely useful when we want to ask many questions
that are largely the same, and differ only in a few assertions:
we can perform the majority of the work once, and then use
\Verb"push" and \Verb"pop" to just assert the differences.
In our small example, the work for the first assertion shared in
both calls to \Verb"check-sat".

@@

The constraint solver has two central pieces of state: the {\em work queue},
and the {\em inert set}.  The work queue contains constraints that need to be
processed, while the inert set contains constraints that have already
been processed.  Constraints are removed---one at a time---from the work queue,
and {\em interacted} with the solver's state.  In the process of interaction
we may solve a wanted constraint, possibly adding new constraints to the
work queue, or notice that a constraint is impossible to solve and record
an error, or---if nothing interesting happens---we add the constraint
to the inert set.  It is also possible that during interaction a previously
inert constraint may be reactivated in a rewritten form and re-inserted
in the work queue.  A single invocation of the constraint solver keeps
interacting constraints until the work queue is empty and all constraints
are in the inert set.

\subsection{Type-Checker Plugins}
Integrating an SMT solver with GHC's type-checker is not the only extension
one might want to have---indeed, there is existing work that extends GHC
with support for units of measure \cite{units-of-measure}.  Instead of
having many ad-hoc extensions directly in GHC's constraint solver, we
collaborated to define and implement and API for extending GHC's functionality
via {\em type-checker plug-ins}.  At present, we are aware of two main
users of this API---our work, and the work on units of measure, but we
hope that this mechanism would make it easier for researcher to experiment
with various extensions to GHC's constraint solver.

An interesting question about type-checker plug-ins is: at what point in
GHC's type checker should we hook them in? We considered a few alternatives:
\begin{enumerate}
\item Add a pass to the solver's pipeline, meaning that plug-ins have to
work with constraints one at a time.
\item Add a call to the plug-ins after the constraints solver has reached
an inert state.
\item Hand off implication constraints directly to the plug-ins,
before invoking GHC's constraint solver.
\end{enumerate}

At the end, we decided on option 2.  Option 3 is the most general, as it
would allow for the plug-in to completely override GHC's behavior.  However,
we were more interested in {\em extending} GHC's capabilities rather than
{\em replacing} them, so we opted against it.  Option 1, on the other hand,
has the closest integration with GHC, but since plug-ins need to precess
constraints one at a time, then they often ended up needing some state,
which than had to be stored and managed somewhere, which became rather
complicated.

We chose option 2 as a nice middle ground, which allows us to look at
all constraints at once, but plug-ins can assume that standard stuff
the GHC knows about has already been eliminated. So, after the constraint
solver reaches an inert state, it calls into the plug-ins, which examine
the inert state and attempt to make progress.  If they do, then the
constraint solver is restarted and process repeats.

\subsection{Improvement and Derived Constraints}
An improving substitution \cite{improvement} allows us to instantiate variables
in the constraints with types, potentially enabling further constraint
simplification.  Of course, we should only do so, as long
as we preserve soundness and completeness.  In this context preserving
soundness means that the new constraints should imply the original
constraints (i.e., we didn't just drop some constraints),
while completeness means that the original constraints imply the new ones
(i.e., we didn't loose generality by making arbitrary assumptions).

In GHC, improvement happens when we rewrite constraints using
{\em equality constraints}. There are three sources of equality constraints:
\begin{itemize}
\item {\em given equalities} are implied by the given constraints,
\item {\em wanted equalities} arise when solving wanted constraints,
\item {\em derived equalities} are implied by the given and wanted constraints,
together.
\end{itemize}

@@

We describe the algorithm in the context of the theory of linear arithmetic
over natural numbers, as having a concrete theory makes it easier to explain
the process and illustrate it with examples.  In the next Section,
we discuss how, and why, we might want to consider other theories also.

\paragraph{Input.} Currently, the algorithm is implemented as a type-checker
plug-in, and so the input to the algorithm is a collection of constraints
that GHC has determine to be inert: GHC simplified everything as much
as it could, improved using equalities, and basically cannot see anything
else to do.  The inert constraints are presented to the plug-in in three
collections, based on the constraints' provenance: one group contains
the given constraints (i.e., assumption), one group contains derived
constraints (see Section~\ref{GHC} for details), and one group contains
the wanted constraints, which are the goals that need solving.

\paragraph{Output.} The desired output of the algorithm is as follows:
\begin{itemize}
  \item Solve as many wanted constraints as possible.
  \item Notice if the wanted constraints are inconsistent.
  \item Compute new given and derived equalities to help solve constraints
        that are outside this decision procedure's scope.
\end{itemize}
The first bullet is clearly the main purpose of the algorithm, but the other
two are quite important also.

The second bullet is useful because it helps us avoid inferring type
schemes with unsatisfiable constraints.  Suppose, for example, that
we had just a single wanted constraint: \Verb"(x + 5) ~ 2".  Since we
are working with natural numbers, this constraint has no solution, so
we will not be able to solve it. As a result, we may end up inferring
a type like this:
\begin{Verbatim}
someFun :: forall x. (x + 5) ~ 2 => ...
\end{Verbatim}
While, technically, this is not wrong, it is undesirable because the type
error is clearly in the definition of \Verb"someFun", but we would delay
reporting the error until the function is called.  So, it is better if
we noticed that the wanted constraints are not satisfiable, are report
an error immediately.

The importance of the third bullet is that it enables collaboration between
the plug-in and the rest of GHC (i.e., the main constraint solver,
and other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd have to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  This equality will then be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3",
which in turn can be discharged by the custom solver for the \Verb"KnownNat"
class.  Such collaboration between different solvers is quite common.
Interestingly, this looks a lot like the collaboration between decision
procedures in an SMT solver!  We discuss this observation further
in Section~\ref{modular-typechecker}.

\subsection{Identifying Relevant Constraints}
Our first task is to identify constraints that are relevant to our solver.
In the current implementation, we consider equalities and inequalities
between a subset of the types of kind \Verb"Nat".   The kind \Verb"Nat" is
inhabited by an infinite family of type-constants:
\begin{Verbatim}
0, 1, 2, .. :: Nat
\end{Verbatim}
These constants may be combined and compared using the following
type-level functions:
\begin{Verbatim}
type family (+)   :: Nat -> Nat -> Nat
type family (*)   :: Nat -> Nat -> Nat
type family (<=?) :: Nat -> Nat -> Bool
\end{Verbatim}
These functions have no user-specified definitions: instead, we extended
the core GHC simplifier with support for forward evaluation on concrete
values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
This simple forward evaluation simplifies and cleans up the constraints,
leaving the more complex reasoning involving variables to the algorithm
being presently described.

The declaration for \Verb"(<=?)" refers to the kind \Verb"Bool", which
is simply the lifted \Verb"Bool" type.  As expected, it is inhabited by the
empty types \Verb"True" and \Verb"False".

So the language of interest to us it constraints formed with the two
predicate symbols, \Verb"(~)" and \Verb"(<=?)", and types made out
of variables, natural number constants, and the operations \Verb"(+)" and
\Verb"(*)".  Furthermore, since we are restricted to working with {\em linear}
arithmetic, we only consider multiplication by a constant.

Of course, constraints may contain other symbols too.  For example,
programmers are free to define their own type functions that return
results of kind \Verb"Nat", so it is entirely possible to encounter
constraints like \Verb"(F 3 + 4) ~ 7", where \Verb"F" is some programmer-defined
type-family.

The process of identifying relevant constraints proceeds as follows:
\begin{enumerate}
\item Eliminate constraints that do not have a relevant top-level predicate
symbol.  The relevant predicate symbols are \Verb"(<=?)", and \Verb"(~)" when
used at kind \Verb"Nat".
\item Import the parameters to the predicates guided by the kind:
  \begin{itemize}
  \item The two supported kinds are \Verb"Nat" and \Verb"Bool"
  \item Currently, there are no interesting operations at kind \Verb"Bool"
  \item If we encounter a type outside of our theory, we name it, and replace
        it with a temporary variable.
  \item We remember the name assignments, so that if the same ``foreign'' type
        appears multiple times, it will get the same name.
  \end{itemize}
\end{enumerate}

The step of naming foreign types is similar to the flattening step performed
by GHC's constraint solver, the main difference being that GHC names every
call to a type-function, while we name only terms outside of the theory.
Still, this duplication of work is somewhat unsatisfactory, and it would
be nice to extend GHC's flattening pass so that it can be reused by
external plug-ins.

\subsection{Consistency and Improvement}

\subsection{Simplifying Constraints}

Given an implication constraint: $G\implies W$.
\begin{enumerate}
\item Identify constraints in the theory
\item Name sub-terms that are outside the theory
\item Assume $G$
\item Check for satisfiability
  \begin{itemize}
  \item If unsat, implication is trivial, record error
  \item If sat, use model for improvement
  \end{itemize}
\item ...
\end{enumerate}

@@

\paragraph{Integers.}
\paragraph{Boolean.}  SAT solver in the type system
\paragraph{Bit-vectors.}  Aribtrary modulo arithmeitc; finite sets of effects.
\paragraph{Sets.}

>>>>

\documentclass{sigplanconf}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fancyvrb}
\usepackage{url}

\newtheorem*{lemma}{Lemma}
\newtheorem*{corollary}{Corollary}

@@

\begin{abstract}
We present a technique for integrating GHC's type-checker with the
an SMT solver.  The technique was developed to add support for reasoning
about type-level functions on natural numbers, and so our implementation
uses the theory of linear arithmetic.  The technique, however, is more
general and makes it possible to experiment with other external decision
procedures, such as reasoning about type-level booleans, bit-vectors,
or any other theory supported by SMT solvers.
\end{abstract}

@@

\section {SMT Solvers}
\label{smt}

@@

This example asks the solver two questions, and the answer to the first
one is \Verb"unsatisfiable", while the answer to the second one is
\Verb"satisfiable" and the solution is \Verb"x = -2".

The \Verb"push" command instructs the solver to save its state, then
proceed as normal.  When the solver encounters a \Verb"pop" command,
it reverts to the last saved state.

This functionality is extremely useful when we want to ask many questions
that are largely the same, and differ only in a few assertions:
we can perform the majority of the work once, and then use
\Verb"push" and \Verb"pop" to just assert the differences.
In our small example, the work for the first assertion is shared in
both calls to \Verb"check-sat".

@@

The constraint solver has two central pieces of state: the {\em work queue},
and the {\em inert set}.  The work queue contains constraints that need to be
processed, while the inert set contains constraints that have already
been processed.  Constraints are removed---one at a time---from the work queue,
and {\em interacted} with the solver's state.  In the process of interaction
we may solve constraint, generate new work, or report impossible constraints.
If nothing interesting happens, then we relocate the constraint to the
inert set.  It is also possible that during interaction a previously
inert constraint may be reactivated in a rewritten form and re-inserted
in the work queue.  A single invocation of the constraint solver keeps
interacting constraints until the work queue is empty and all constraints
become inert.

\subsection{Type-Checker Plugins}
In this paper we describing a fairly general extension to the constraint
solver, however one could always imagine others too. Indeed, there is
existing work that extends GHC with support for units of measure
\cite{units-of-measure}.  Instead of having many ad-hoc extensions directly
in GHC's constraint solver, we collaborated to define and implement an
API for extending GHC's functionality via {\em type-checker plug-ins}.
At present, we are aware of two main
users of this API---our work, and the work on units of measure. We
hope, however, that this mechanism would be useful to other researchers too,
as it makes it fairly easy to experiment with various extensions to
GHC's constraint solver.

An interesting question about type-checker plug-ins is: at what point in
GHC's type checker should we invoke them?  The answer to this question
affects their functionality, and the interface that a plug-in would have
to implement.  We considered a few alternatives:
\begin{enumerate}
\item Add a pass to the solver's pipeline, meaning that plug-ins process
constraints one at a time, the way GHC does.
\item Add a call to the plug-ins after the constraints solver has reached
an inert state.
\item Hand off implication constraints directly to the plug-ins,
before invoking GHC's constraint solver.
\end{enumerate}

Option 1 has the closest integration with GHC, but since plug-ins need
to process constraints one at a time, then they often ended up needing
some state, which than has to be stored and managed somewhere,
which became rather complicated.

Option 3 is the most general, as it
would allow for the plug-in to completely override GHC's behavior.  However,
we were more interested in {\em extending} GHC's capabilities rather than
{\em replacing} them, so we opted against it.

We chose option 2 as a nice middle ground. It allows a plug-in to
work with all constraints at once, but it has the benefit that standard work
done by GHC has already happened. So, after the constraint
solver reaches an inert state, it calls into the plug-ins, which examine
the inert state and attempt to make progress.  If they do, then the
constraint solver is restarted and process repeats.

\subsection{Improvement and Derived Constraints}
An improving substitution \cite{improvement} allows us to instantiate variables
in the constraints with types, potentially enabling further constraint
simplification.  Of course, we should only do so, as long
as we preserve soundness and completeness.  In this context, preserving
soundness means that the improved constraints should imply the original
constraints (i.e., we didn't just drop some constraints),
while completeness means that the original constraints imply the improved ones
(i.e., we didn't loose generality by making arbitrary assumptions).

In GHC, improvement happens by rewriting with {\em equality constraints}.
There are three sources of equality constraints:
\begin{itemize}
\item {\em given equalities} are implied by the given constraints,
\item {\em wanted equalities} arise when solving wanted constraints,
\item {\em derived equalities} are implied by the given and wanted constraints,
together.
\end{itemize}

@@

We describe the algorithm in the context of the theory of linear arithmetic
over the natural numbers, as having a concrete theory makes it easier to explain
the process and illustrate it with examples.  In the next Section,
we discuss how, and why, we might want to consider other theories also.

\paragraph{Input.} Currently, the algorithm is implemented as a type-checker
plug-in, and so the input to the algorithm is a collection of constraints
that GHC has determine to be inert. This means that GHC simplified everything
as much as it could, improved using equalities, and basically cannot see
anything else to do.  The inert constraints are presented to the plug-in
in three collections, based on the constraints' provenance: one group contains
the given constraints (i.e., assumptions), one group contains derived
constraints (see Section~\ref{GHC} for details), and one group contains
the wanted constraints, which are the goals that need solving.

\paragraph{Output.} The desired output of the algorithm is as follows:
\begin{itemize}
  \item Solve as many wanted constraints as possible.
  \item Notice if the wanted constraints are inconsistent.
  \item Compute new given and derived equalities to help solve constraints
        that are outside this decision procedure's scope.
\end{itemize}
The first task is the most obvious purpose of the algorithm, but the other
two are quite important also.

Noticing inconsistencies avoids the inference of types with unsatisfiable
constraints.  Consider, for example, a single wanted constraint,
\Verb"(x + 5) ~ 2".  Since we are working with natural numbers,
this constraint has no solution, so we cannot solve it.
As a result, we may end up inferring a type like this:
\begin{Verbatim}
someFun :: forall x. (x + 5) ~ 2 => ...
\end{Verbatim}
While, technically, this is not wrong, it is undesirable because the type
error is clearly in the definition of \Verb"someFun", but we would delay
reporting the error until the function is called.  So, we'd like to
notice constraints that are impossible to solve (i.e., they are logically
equivalent to $\bot$), so that we can report type-errors that are closer
to their true location.  Note that the source of a contradiction may
be a combination of constraints, and not just a single one.  For example,
the constraints \Verb"x >= 5" and \Verb"3 >= x" are inconsistent together,
but have solutions when considered individually.

Computing new equations is also very important, as it enables collaboration
between the plug-in and the rest of GHC (i.e., the main constraint solver,
and other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd like to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  Then, this equality can be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3"
which, in turn, can be discharged by the custom solver for the \Verb"KnownNat"
class.  Such collaboration between different solvers is quite common.
Interestingly, this looks a lot like the collaboration between decision
procedures in an SMT solver!  We discuss this observation further
in Section~\ref{modular-typechecker}.

% \subsection{Communicating With The Solver}
% Our implementation uses the \Verb"simple-smt" library \cite{simple-smt}
% to communicate with external SMT solvers.

\subsection{The Language of Constraints}
Our first task is to identify constraints that are relevant to our solver.
In the current implementation, we consider equalities and inequalities
between a subset of the types of kind \Verb"Nat".   The kind \Verb"Nat" is
inhabited by an infinite family of type-constants:
\begin{Verbatim}
0, 1, 2, .. :: Nat
\end{Verbatim}
These constants may be combined and compared using the following
type-level functions:
\begin{Verbatim}
type family (+)   :: Nat -> Nat -> Nat
type family (*)   :: Nat -> Nat -> Nat
type family (<=?) :: Nat -> Nat -> Bool
\end{Verbatim}
These functions have no user-specified definitions: instead, we extended
the core GHC simplifier with support for forward evaluation on concrete
values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
This simple forward evaluation cleans up the constraints, leaving the more
complex reasoning---involving variables---to the algorithm being presently
described.

The declaration for \Verb"(<=?)" refers to the kind \Verb"Bool", which
is simply the lifted \Verb"Bool" type.  As expected, it is inhabited by the
empty types \Verb"True" and \Verb"False".   Having \Verb"(<=?)" be a function
that returns a boolean is a little more convenient for programmers than having
a constraint. In this form, a programmer may inspect the result of the
inequality and make a decision based on it.  We can use a simple
abbreviation to implement the corresponding inequality constraint:
\begin{Verbatim}
type (<=) a b  = (a <=? b) ~ True
\end{Verbatim}

The constraints of interest to the algorithm are summarized by the following
grammar:
\begin{align*}
c & = e \overset {\mathbb{N}} \sim e ~|~  e \overset {\mathbb{B}} \sim e \\
e^\mathbb{N} & = \alpha ~|~ n ~|~ e + e ~|~ n * e \\
e^\mathbb{B} & = \alpha ~|~ \mathrm{False} ~|~ \mathrm{True} ~|~ e \leq^? ~e \\
n          & = 0 ~|~ 1 ~|~ \dots
\end{align*}

\subsection{Importing Constraints}

Of course, general constraints may be more complex.  In particular,
programmers are free to define their own type functions that return
results of kind \Verb"Nat", so it is entirely possible to encounter
constraints like \Verb"(F 3 + 4) ~ 7", where \Verb"F" is some programmer-defined
type-family.  It is also possible to encounter non-linear constraints such
as \Verb"x * y".

The process of identifying relevant constraints is as follows:
\begin{enumerate}
\item Set aside constraints that do not have a relevant top-level predicate
symbol.
\item Import the parameters to the predicates guided by the kind.
\item If we encounter a type outside of our theory, then we name it and replace
      it with a variable.
\item We remember the names assigned to various types, so that we may reuse
      the same name if a type appears multiple times.
\end{enumerate}

For example, the constraint \Verb"(F 3 + 4) ~ 7" will be imported as
\Verb"(x + 4) ~ 7", and we will remember that \Verb"x" stands for \Verb"F 3".
Later on, if we need to return results to the core GHC solver that mention
\Verb"x", we replace it by the orignal expression, \Verb"F 3".

The process of abstracting foreign constraints makes sense in our context,
because it {\em generalizes} the constraint.  Consider a constraint $C(t)$, and
its generalized form $C(x)$, where the sub-term $t$ is replaced by
the variable $x$, which does not occur in $C(t)$.

\begin{lemma}
If $C(t)$ is satisfiable, then $C(x)$ is also satisfiable.
\begin{proof} This follows because, by definition,
$C(t) \iff C(x) \wedge x = t$. Therefore, if $\sigma$ is satisfying
assignment for $C(t)$, then $\sigma \cup \{ x = \sigma t \}$ is satisfying
assignment for $C(x)$.
\end{proof}
\end{lemma}

\begin{corollary} If $C(x)$ is not satisfiable, then neither is $C(t)$.
This is simply the contrapositive form of the previous Lemma.
\end{corollary}

Recall from Section~\ref{smt}, than to prove something, we make sure
that there are no counter examples.  The Corollary states that if there
are no counter examples to the abstracted constraint, than there will
be no counter-examples to the original constraint as well, and so our
proof is sound.  From a logical stand-point, naming away sub-terms
amounts to trying to prove a more general fact, and so if we can complete
the proof in the more general setting, then the proof will work for
the special case of the original constraint.

\paragraph{Aside.}
The step of naming foreign types is similar to the flattening step performed
by GHC's constraint solver, the main difference being that GHC names every
call to a type-function, while we name only terms outside of the theory.
Still, this duplication of work is somewhat unsatisfactory, and it would
be nice to extend GHC's flattening pass so that it can be reused by
external plug-ins.

\subsection{Checking for Consistency}

Given some constraints in our theory, we would like to know if a
solution exists.  If the constraints are inconsistent, then we can
terminate the constraint solving problem early, and report an error,
as discussed previously.

To check for consistency, we assert all constraints, and ask the SMT
solver if the result is satisfiable.  If this is not the case, then
we report and error, otherwise we proceed with the algorithm.  Note that
since we are working with generalized constraints, if we find an
inconsistency, then we are sure that we've detected a problem.  However,
if the SMT solver says that the constraints are satisfiable, then
we now that this is the case from the point of view of our theory,
however the constraints might still be unsatisfiable if one took a global
view of the problem.  Consider, for example, \Verb"(F a + 4) ~ 7".
This will get imported into our plug-in as \Verb"(x + 4) ~ 7", which is
satisfiable by \Verb"x = 3".  Now, some other theory might know that
\Verb"F a ~ 3" is not actually possible, but we won't detect this problem
at this stage.

\subsection{Minimizing Conflicts}
If we detect a contradiction, then
we know that there is no possible solution to the constraints.  However,
often only a few of the constraints are involved in the actual problem,
and it is much nicer if we report only them as the cause of the error,
rather than getting a huge error involving all constraints.  If an SMT
solver detects that a collection of assertions is inconsistent, it may
be able to compute an {\em unsatisfiability core}, which is the sub-set
of the assertions that lead to the conflict---exactly what we want!

Unfortunately, not all SMT solvers support this feature and CVC4---%
the SMT solver that we used during the development of this work---%
does not have this ability.  To work around this, we implemented
a custom minimization algorithm which work as follows:
\begin{Verbatim}
minimize s yes [] = return yes
minimize s yes mb =
  do solverPush s
     search s yes [] mb

search s yes mb (c : cs) =
  do solverAssert s c
     res <- solverCheck s
     case res of
       Unsat ->
         do solverPop s
            solverAssert s c
            minimize s (c : yes) mb

       _ -> search s yes (c : mb) cs
\end{Verbatim}
The algorithm keeps track of constraints that we are certain are part
of the conflict in the variable \Verb"yes".  In addition, the algorithm
has the invariant that the constraints in \Verb"yes" are always asserted
in the solver's state, represented by the variable \Verb"s".
Initially, \Verb"yes" starts off empty.

We search for conflicts in the function \Verb"search", which asserts
constraints one at a time, until a conflict is discovered.  The constraint
that caused the conflict is added to the \Verb"yes" set, and then we
examine the previously asserted constraint, to check if they are necessary
for the conflict.

The minimization process performs in $O(n^2)$ time, where $n$ is the
number of constraints.  This has not been a big problem,
as this minimization happens only when we report errors and, typically,
the number of constraints is fairly low.

\subsection{Improvement}

If all constraint are consistent (i.e., have at least one solution),
then we check for computed equality constraints.  As discussed previously,
these help other parts of the constraint solver to make progress.

It is convenient to compute the improvements right after the consistency
check because at this point we already have all constraints asserted
in the solver's state.  If all constraints in scope are given constraints,
then we also generate given equalities, otherwise we generate derived
equalities.

The current implementation considers three forms of improvement,
two of which are completely generic, and one of which is specific to
linear arithmetic.

\paragraph{Improve to Constant.} As we know that the constraints are
consistent, we can ask the SMT solver for a satisfying assignment.
This assignment contains one possible solution to constraints, but
we should not emit an improving equality unless we are sure that
this is the {\em the only} possible solution.  So, if $x = v$ is
in the satisfying assumption, then we try to prove that the currently
asserted constraints imply this fact.  We do this by temporarily asserting
$x \neq v$, and checking for satisfiability.  If the resulting is unsatisfiable,
then we know that $v$ is the only possible value for $x$,
and we can emit the corresponding improving equation.
\begin{Verbatim}
solverProve s p =
  do solverPush s
     solverAssert s (SMT.not p)
     res <- solverCheck s
     solverPop s
     return (res == Unsat)

mustBeK s x v =
  solverPorve s (eq (SMT.const x) (value v))
\end{Verbatim}
This process takes $O(n)$ time, where $n$ is the number of variables.
Note that here we are making use of the incremental capabilities of the
solver, and reusing all asserted constraints, just doing one
additional \Verb"assert" per variable. To see the process in action,
consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
are:
\begin{Verbatim}
(assert (= (* 2 x) 16))
(check-sat)
; SMT response: SAT
(get-value (x))
; SMT response: (8)
(push 1)
(assert (not (= x 8)))
(check-sat)
; SMT response: UNSAT
(pop 1)
; We generate a computed improvement: x ~ 8
\end{Verbatim}

\paragraph{Improve to Variable.}
We may also use the satisfying assignment to look for improvements of
the form $x = y$, where $x$ and $y$ are both variables.
We look at the satisfying assignment, ignoring variables
that were already improved to constants, and consider pairs of variables
that happen to have the same value in this assignment. If $x$ and $y$
are two such variables, then we try to prove that $x = y$ must hold
under the current assumptions:
\begin{Verbatim}
mustEqual s x y =
  solverProve s (eq (SMT.const x) (SMT.const y))
\end{Verbatim}
This process take $O(n^2)$ time, where $n$ is the number of variables,
but we only need to call the solver for pairs of variables that have the same
value---if they do not, then the current assignment provides a counter
example indicating that the variables do not need to be equal in general.

\paragraph{Improve Using a Linear Relation.}
This improvement is a generalization of the previous example, that
is specific to linear arithmetic.  The idea is to try to discover
improving equations of the form $y = A * x + B$.  This type of
improvement is a little different from the others, in that the right-hand
side of the equation contains terms that are in the theory that we are solving.
In general, such constraints are not very useful for the rest of GHC,
as other parts of the constraint solver do not know about our theory
(i.e., the functions \Verb"*" or \Verb"+") and, if we can prove it, then
we must already know about this improvement!  This sort of improvement, is
useful in one very common case, however:  if $y$ is a unification variable,
then having this type of improving equation enables GHC to instantiate
the variable, which leads to better type inference and simpler constraints,
so we attempt this improvement only when the previous two have failed,
and only for the unification variables in the constraints.

We can compute the values for $A$ and $B$, if we have two examples for
$x$ and two examples for $y$. Note that each call to the solver in
the previous improvement steps gave rise to a potential additional set
of examples, so we have plenty of points to work with: we take care to
remember the values for the variables from two models.   It is also
important that the $x$ values be different, but we can always find such
examples: if we cannot find such an example, then $x$ would have only
one possible value, and it would have been improved already.  It turns
out that this sort of situations occur fairly frequently.  Consider,
for example, the following program fragment:
\begin{Verbatim}
f :: Proxy (a + 1)
f = Proxy

g :: Proxy (b + 2)
g = f
\end{Verbatim}
In this example, GHC needs to infer how to instantiate \Verb"f"'s type
parameter, \Verb"a", and to do so, our plug-in needs to figure out that
\Verb"a" may be instantiate to \Verb"b + 1".

Once we compute candidate values for $A$ and $B$, we still need to
invoke the solver to validate that the compute relation holds in all
instances of the constraints, and not just the two that we happened to pick.

Our current implementation only considers linear relations between pairs
of variables as we were concerned about the performance penalty if
we tried combinations of more variables: the math will work out, but
the algorithm might become too slow to be practical, and alternative
solutions based on rewriting might be more suitable.

\paragraph{Custom Improving Rewrites.}
All the improvements that we've defined so far work by {\em only} using
the solver---note that we were just looking at satisfying assignments,
and did not have any special rules about the shapes of the constraints.
It also makes sense to extend the system with custom rewrite rules,
which has the potential of speeding up performance, and also adding
support for features that go beyond the solver's capabilities.
We have experimented with various custom rules, but it is as yet
unclear what constitutes a good set of rewrites.

\subsection{Solving Constraints}

The process of solving constraints is a straight-forward call the
solver.  The only thing we need to do before solving constraints is
to remove the assertions that were added during the consistency and
improvement stages.  Recall that GHC solves implication constraints
with two call to the constraint solver: the first one asserts
the given constraints, while the second one actually solves goals.

While we are asserting the assumption, we just check for consistency
and improvement (to generate new given equalities), but we perform
no solving, as there is nothing to solve.  During the solving stage,
we mark the solver's state after we've asserted the givens,
then we assert the new goals, check for consistency, and improvement,
and then, before solving, revert back to the state where only
the givens are asserted.
\begin{Verbatim}
solverSimplify s wanteds =
  solverPrepare s wanteds $ \others our_wanteds ->
  do res <- mapM tryToSolve our_wanteds
     let (unsolved, solved) = partitionEithers res
     return (solved, unsolved ++ others)
  where
  tryToSolve (ct,e) =
    do proved <- solverProve s e
       if proved
         then return (Right ct)
         else return (Left ct)
\end{Verbatim}
The call to \verb"solverPrepare" identifies the constraints that belong
to our theory and translates them the SMTLIB language, where \Verb"others"
are the constraints that are completely outside our theory, and
\Verb"out_wanteds" are the constraints that we know about. Then,
we invoke the solver for each goal, and see if we can prove it.

\paragraph{Evidence.} When we solve a constraint, GHC expects some
evidence to be produced, explaining why the constraint holds
\cite{fc-evidence}.  The same is expected when generating new given
constraints.  While this evidence is not used for code generation,
it is a very useful for sanity checking while working with the constraint
solver.  Unfortunately, at present, our plug-in does not produce any
meaningful evidence, beyond indicating that the fact was produce
by using the SMT solver. In principle, SMT solvers should be able to
produce a proof, when they have concluded that a set of assertions
is unsatisfiable.  Unfortunately, this is not a commonly used feature,
so many provers do not support it out of the box, and the format of the
proofs is not standardized.

@@

In the previous Section, we described the core working of a procedure
for integrating an SMT solver with GHC's constraint solver.  While
we concentrated on the theory of natural numbers and linear arithmetic,
little of the algorithm is specific to that theory.

It would be fairly easy to add support for other theories, simply
by declaring additional ``uninterpreted'' kinds and type-functions
in Haskell.  In the rest of this Section we explore some of these
possibilities.

\subsection{A Theory for Booleans}
We already made use of the lifted \Verb"Bool" type, but so far Haskell
does not support any interesting operations, beyond the \Verb"<=?" relation
that we described.  Programmers are already experimenting with defining
type functions. For example, with the current version of GHC we may
define a function for conjunction of type-level booleans:
\begin{Verbatim}
type family And a b where
  And False x = False
  And True x  = x
\end{Verbatim}

Definitions like this work for simple evaluation, but do not support
more sophisticated reasoning.  For example, if GHC encounters
the constraint \Verb"And x y ~ True", it will not be able to conclude
that the only way to solve this is if \Verb"x ~ True" and \Verb"y ~ True"
both hold.  Using the approach outlined in the previous Section, we
gain the full power of a SAT solver in the type system---SMT solvers
can reason about booleans, indeed a SAT solver is usually an important
part of the implementation of an SMT solver.

A good starting point for the theory of booleans, would be the following
signature:
\begin{Verbatim}
type family And (a :: Bool) (b :: Bool) :: Bool
type family Or  (a :: Bool) (b :: Bool) :: Bool
type family Not (a :: Bool)             :: Bool
\end{Verbatim}

\subsection{A Theory for Integers}

So far, our work uses natural numbers, which are quite useful for
keeping track of the sizes of data-structures such as arrays, vectors,
lists, bit-vectors, etc.  The standard theory supported by the SMT solvers
is actually linear arithmetic over the integers---in our implementation
we have to do some extra work to assert that we are only interested
in non-negative solutions.

The main challenge for supporting both integers and natural numbers
at the type level is not the technology for solving the equations, but
the notation!  One possible signature for the theory of type-level
integers is:
\begin{Verbatim}
type family ToInt  (a :: Nat) :: Int
type family Negate (a :: Int) :: Int
type family Add    (a :: Int) (b :: Int) :: Int
type family Sub    (a :: Int) (b :: Int) :: Int
type family Mul    (a :: Int) (b :: Int) :: Int
\end{Verbatim}
While this notation is somewhat verbose, at least it has the benefit
of exposing the functionality to programmers, who may use a more suitable
notation depending on their needs.  For example, if a program works
mostly with integers, one could add the following declaration, to get
the more usual mathematical notation:
\begin{Verbatim}
type (+) a b = Add a b
\end{Verbatim}

\subsection{A Theory of Bit-Vectors}
SMT solvers also have good support for reasoning about fixed-length bit-vectors.
Unlike the theory of linear arithmetic, the theory of bit-vectors
supports arbitrary multiplication and division, as well as various bit-wise
operations, similar to the operations available in hardware.

\paragraph{Tracking Sizes.}
If we are interested in keeping track of the sizes of data-structures
in memory, using 64-bit bit-vectors is a reasonable compromise
which provides more expressive reasoning than linear arithmetic,
and accommodates virtually all data-structures that fit in a computer's memory.

\paragraph{Finite Sets.}
Another potential application for the theory of bit-vectors is
to implement finite sets of integers at the type level.
As is usual at the value level, each bit is used to indicating the presence
or absence of an element in the set, and the bitwise \Verb"and"
and \Verb"or" operations correspond to intersection and union.
Such functionality would be useful in the implementation of various effect
systems.
